---
title: "DAND - Exploratory Data Analysis Project"
author: "Guillermo Naranjo"
date: "October 2, 2017"
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Loading used packages.
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(scales)
library(GGally)
library(memisc)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}

# Loading the data and reorganizing dataframe
extract_df <- function(df){

  df <- read.csv('wineQualityReds.csv')
  
  #I use quality level factor to show graphs by color level
  df['quality_level'] <- as.factor(df$quality)

  #Since quality level 3 and 8 have few observations I've grouped quality levels
  #to increase significance and apply further analysis:
  #Quality1 = 3 + 4, 
  #Quality2 = 5
  #Quality3 = 6
  #Quality4 = 7 + 8
  quality_groups <- c(1,1,2,3,4,4)
  df['quality_group'] <- as.factor(quality_groups[df$quality_level])
  
  df$X <- NULL
  return(df)
}

pf <- extract_df()

```

```{r Utilitary functions, echo=FALSE}

#Outliers detection function used to explore outlier effect in analysis
olimits <- function(variable) { 
  fivenum(variable)
  minoutlier <- quantile(variable,0.25)-1.5*IQR(variable)
  maxoutlier <- quantile(variable,0.75)+1.5*IQR(variable) 
  return(c(minoutlier,maxoutlier))
} 

#Graph used to show Univariate histogram plots 
show_histo <- function(df, var, label, bwidth){
    g <- ggplot(aes(x = var), data = df) +
    scale_y_sqrt()+
    geom_histogram(binwidth = bwidth, color = 'darkgrey', fill = 'grey') + 
    theme(legend.position="bottom") + 
    xlab(label) +
    ylab('Number of observations')
  return(g)
}

#Graph used to show Univariate histogram plots 
show_histo_no_outliers <- function(df, var, label, bwidth){
    o <- ggplot(aes(x = var), data = df) +
    scale_y_sqrt()+
    coord_cartesian(xlim = olimits(var)) +
    geom_histogram(binwidth = bwidth, color = 'darkgrey', fill = 'grey') + 
    theme(legend.position="bottom") + 
    xlab(label) +
    ylab('Number of observations')
  return(o)
}

#Graph used to show Univariate box plots 
show_box <- function(df, var, label){
  b <- ggplot(df, aes(y = var, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = var), alpha = 0.2) +
  labs(y = label, x = '',title = "") +
  coord_flip()
  
  return(b)
}

#Graph used to show bivariate violin plots 
show_violin <- function(df, var, label){
  s <- ggplot(df, aes(x=quality_level, y=var)) + 
  geom_violin(trim=FALSE)+
  geom_boxplot(width=0.1, fill="white")+
  labs(x="quality", y = label)
  s<- s + theme(legend.position="none")
  return(s)
}
```

# DataSet:

I'm using red wine dataset since I'm particularly interested in classification
techniques and I also like wine ... good deal. As mentioned in the [README file](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt),

# Univariate Plots Section

```{r VariablesSummary, echo=FALSE, cache=FALSE}
str(pf)
```

> Redwine quality varible contains six leves, two of them (3 low quality wines
and 8 high quality wines) have few cases. I will not apply test significance of
those levels but to have a better understanding for colors I group: 

* level 3 and 4 as Cat1
* levels 7 and 8 as Cat4.



#### Acidity / basis variables

The following grid shows the distribution of acidity/basis variables.

```{r Distributions1, echo=FALSE}
h1 <- show_histo(pf,pf$fixed.acidity,'Fixed Acidity',0.5)
h2 <- show_histo(pf,pf$volatile.acidity,'Volatile Acidity',0.05)
h3 <- show_histo(pf,pf$citric.acid,'Citric Acid',0.05)
h4 <- show_histo(pf,pf$pH,'pH',0.05)
grid.arrange(h1,h2,h3,h4, ncol=2)
```

fixed.acidity
```{r echo=FALSE, cache=FALSE}
summary(pf$fixed.acidity)
```

volatile.acidity
```{r echo=FALSE, cache=FALSE}
summary(pf$volatile.acidity)
```

citric.acid
```{r echo=FALSE, cache=FALSE}
summary(pf$citric.acid)
```

pH
```{r echo=FALSE, cache=FALSE}
summary(pf$pH)
```


* Fixed acidity is unimodal, right skewed with peak at 7.9.
* Volatile acidity is multimodal, it would be interesting to see in bivariate
analysis if those modals are related to quality or other variables.
* Citric acid like volatile acidity is multimodal and looks complex.
* pH seems to be a normal distribution with peak at 3.310.

> **Courious**: are modals related to quality?

#### Alcohol and Sulphates

The following grid shows the distribution of alcohol and sulphated variables.

```{r Distributions2, echo=FALSE}
h5 <- show_histo(pf,pf$free.sulfur.dioxide,'Free Sulfur Dioxide',3)
h6 <- show_histo(pf,pf$total.sulfur.dioxide,'Total Sulfur Dioxide',12)
h7 <- show_histo(pf,pf$sulphates,'Sulphates',0.07)
h8 <- show_histo(pf,pf$alcohol,'Alcohol',0.28)
grid.arrange(h5,h6,h7,h8, ncol=2)
```

free.sulfur.dioxide
```{r echo=FALSE, cache=FALSE}
summary(pf$free.sulfur.dioxide)
```

total.sulfur.dioxide
```{r echo=FALSE, cache=FALSE}
summary(pf$total.sulfur.dioxide)
```

sulphates
```{r echo=FALSE, cache=FALSE}
summary(pf$sulphates)
```

alcohol
```{r echo=FALSE, cache=FALSE}
summary(pf$alcohol)
```


* Free sulfur dioxide is unimodal with peak at 14.00, completely right skewed
wth some outlier passing 60g/dm3.
* Total sulfur dioxide distribution is similar to Free sulfur dioxide with peak
at 38.00 and outliear passing 200ppm TSD.
* Suplates, unimodal, not completely normal with peak at 0.6200 g/dm3 (a log
transformation on X will help) with some possible outliers passing 1.5 g/dm3.
* Alcohol, biimodal with higuest peak at 10.20% and second one at 11%.

#### Other less interesting variables


```{r Distributions3, echo=FALSE}
h9 <- show_histo(pf,pf$residual.sugar,'Residual Sugar',1)
h10 <- show_histo(pf,pf$density,'Density',0.0008)
h11 <- show_histo(pf,pf$chlorides,'Chlorides',0.03)
grid.arrange(h9,h10,h11,h8,ncol=2)
```

residual.sugar
```{r echo=FALSE, cache=FALSE}
summary(pf$residual.sugar)
```

density
```{r echo=FALSE, cache=FALSE}
summary(pf$density)
```

chlorides
```{r echo=FALSE, cache=FALSE}
summary(pf$chlorides)
```

* If we remove some outlier from Chlorides and Residual Sugar distributions
they look unimodal, with some "special cases" at higher sugar/chlorides level.
* Density looks completely normal distribuided with peak at 0.9968.


#### Boxplots

Now, lets have a closer looks at Residual sugar, Chlorides and Sulphates to see
how outliers affect each variable.

```{r Boxplots2, echo=FALSE}
b1 <- show_box(pf,pf$residual.sugar,'Residual sugar')
b2 <- show_box(pf,pf$chlorides,'Chlorides')
b3 <- show_box(pf,pf$sulphates, 'Sulphates')
grid.arrange(b1,b2,b3,ncol=2)
```

It seems both residual sugar and chlorides have some outliers that could affect
the distribution. So, there are few cases where wine is either sweet or salty.

> outlier detections is applied.

```{r Boxplots3, echo=FALSE}
o1 <- show_histo_no_outliers(pf,pf$residual.sugar,'Residual sugar',0.2)
o2 <- show_histo_no_outliers(pf,pf$chlorides,'Chlorides',0.005)
o3 <- show_histo_no_outliers(pf,pf$sulphates, 'Sulphates',0.05)
grid.arrange(o1,o2,o3,ncol=2)
```

# Univariate Analysis

**What is the structure of your dataset?**

Our dataset contains 1599 observations and 11 quantitative variables.
This is a tidy data set that describes each observation (red variants of the
Portuguese "Vinho Verde" wine) with 11 qualitative  physicochemical input
variables and 1 output variable: quality. I'll try to analyze which variables
define red wine quality for this particular sample. There are some interesting
facts about variables that could relate to quality: 

* Volatile acidity: at too high of levels can lead to unpleasant, vinegar taste.
* Citric acid: can add 'freshness' and flavor to wines.
* Free sulfur dioxide: prevents microbial growth and the oxidation of wine (SO2).
* Total sulfur dioxide: over 50 ppm becomes evident in the nose and taste.
* Density: dependent variable of alcohol and sugar.
* pH: how acid or basic is the wine, most wines are between 3-4 on the pH scale.
* Sulphates: additive which can contribute to S02 levels.
* Alcohol: the percent alcohol content of the wine.

**What is/are the main feature(s) of interest in your dataset?**

Volatile acidity, citric acid, sulphates and alcohol are interesting since
they are multimodal. Is there a relation between each mode and quality?

**What other features in the dataset do you think will help support your**
**investigation into your feature(s) of interest? **

Since Fixed acidity and pH are related to citric and volatile acidity those are
other variables I would like to analyze. As for alcohol, it looks like there is
a relation to density and sugar residual that we need to verify.

**Did you create any new variables from existing variables in the dataset?**

Not yet but Density, pH and Fixed acidity can eventually be removed if we
managed to conclude they can be highly correlated to other varibles linearly
related to quality.

**Of the features you investigated, were there any unusual distributions?**

Sugar residuals, chlorides, sulphates and total sulfur dioxide look normal once
outliers are applied. Residual Sugar and Sulphates are even alike.

Some are normal distributions like Ph and Density but in most cases the
distributions are right skewed. I applied a **scale_y_sqrt** to the plots
to amplify the modals and see other patterns.

# Bivariate Plots Section

Since our Redwine data set contains only continuous variables we are able to
apply Pearson correlation.

```{r, echo=FALSE}
corr_pf <- pf
corr_pf$quality_level <- NULL
corr_pf$quality_group <- NULL
colnames(corr_pf) <- c("F.Acid", "V.Acid", "C.Acid","RSugar","Chlor","FreeSD", "TotalSD","Dens.","pH", "Sulph","Alcoh","Qual.")
res <- cor(corr_pf)
round(res, 2)
```

Besides Free Sulfate Dioxide and Free Sulfate Dioxide (that are closely related),
lets look at other interestingly related variables:

```{r, echo=FALSE}
corr_pf$Qual <- pf$quality_group
ggpairs(corr_pf, bins = 30, columns = c("F.Acid","V.Acid", "C.Acid","Dens.","pH"),
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

```{r, echo=FALSE}
ggpairs(corr_pf, bins = 30, columns = c("RSugar","Dens.", "Chlor","Sulph","Alcoh"),
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

And finally lets plot the the variables more correlated to quality with 
conditional summaries:

```{r, echo=FALSE}
quality_groups <- group_by(pf, quality)
val_by_quality <- summarise(quality_groups,
          fa_Median = median(fixed.acidity),
          va_Median = median(volatile.acidity),
          ca_Median = median(citric.acid),
          rs_Median = median(residual.sugar),
          ch_Median = median(chlorides),
          fs_dMedian = median(free.sulfur.dioxide),
          ts_dMedian = median(total.sulfur.dioxide),
          de_Median = median(density),
          pH_Median = median(pH),
          su_Median = median(sulphates),
          al_Median = median(alcohol),
          n = n())
t(round(val_by_quality, 3))
```

```{r Violinplots, echo=FALSE}
s1 <- show_violin(pf,pf$volatile.acidity,'Volatile acidity')
s2 <- show_violin(pf,pf$chlorides,'Chlorides')
s3 <- show_violin(pf,pf$sulphates, 'Sulphates')
s4 <- show_violin(pf,pf$alcohol, 'Alcohol')
grid.arrange(s1,s2,s3,s4,ncol=2)

```

**Observations:**

* Fixed Acidity mean increases as quality increases with a peak at 8.8 g/dm^3.
Almost logistic but with more variance at quality level 8.
* **Volatile acidity decreases as quality increases and has less variance** 
**as quality increases as well**.
* Citric acid follow a positive linear behaviour as quality increases.
* Residual Sugar doesn't have a linear behaviour with quality.
* **Chlorides (once ouliers are excluded) seems to follow a linear model** against 
quality by looking at conditional summary.
* The same goes for pH and density. There is a relation with quality but not
completely linear.
* Alcohol has a strong correlation to quality **except for level 5** but as the
violin plot shows it shows outliers and particular behaviour compared to other
levels.
* **Sulphates seems to have a linear pattern to quality**.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

We have some strong relations:
* Between variables related to acidity: fixed acidity, volatile acidity, citric
acid and pH, been Volatile acidity the one with higuest correlation to quality.
**I conclude red wines with less volatile acidity levels tend to have high quality**.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Density also have a relation to acidity variables specially to Fixed acidity
(0.668).

It's interest that Density is also related to alcohol (-0.50). So, it seems
density can be a link been acidity variables and alcohol/sulphates variables to
define quality.

### What was the strongest relationship you found?

* pH and Fixed acidity was the strongest with 0.68.
* **Compared to quality, Alcohol with 0.48 was the strongest. Followed by**
**volatile acidity**

# Multivariate Plots Section

> **Quality levels grouping:** to simplify analysis I've collapsed quality
levels from 6 to 4 levels. This will visualize more easily the patterns between
variables that affect quality.

```{r, echo=FALSE}
corr_pf$Qual <- pf$quality_group
ggpairs(corr_pf, bins = 30, columns = c("F.Acid","V.Acid", "C.Acid","Dens.","pH"),
        mapping=ggplot2::aes(colour = Qual))
```

```{r, echo=FALSE}
ggpairs(corr_pf, bins = 30, columns = c("RSugar","Dens.", "Chlor","Sulph","Alcoh"),
        mapping=ggplot2::aes(colour = Qual))
```

**Other interesting variables that relates to quality**

Fixed acidity and density show a positive correlation that translates also into
quality.

```{r, echo=FALSE}
ggplot(aes(x = density, y=fixed.acidity, colour = quality_level), data = pf) +
  geom_smooth(se = FALSE, method = "lm") +
  geom_point(alpha = 1/3, size = 4) +
  scale_color_brewer(type='seq')
```

Alcohol and density show also a correlation that translates into quality

```{r, echo=FALSE}
ggplot(aes(x =density, y=alcohol, colour = quality_level), data = pf) +
  geom_smooth(se = FALSE, method = "lm") +
  geom_point(alpha = 1/3, size = 4) +
  scale_color_brewer(type='seq')
```

# Multivariate Analysis

By selecting the variables that are more related to quality grouped by:
1. Alcoholic (excluding density and suger residuals since are related to alcohol)
2. Acidity (volatile, since others like pH, Citric and Fixed are related as well)
3. Sulphates
3. Chloridic
4. SO2 (i included SO2 since it improved the model a little bit)

This is the model:

```{r}
m1 <- lm(quality ~ alcohol, data = pf)
m2 <- update(m1, ~ . + volatile.acidity)
m3 <- update(m2, ~ . + I(log(sulphates)))
m4 <- update(m3, ~ . + chlorides)
m5 <- update(m4, ~ . + total.sulfur.dioxide)
mtable(m1, m2, m2, m3, m4, m5)
```

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I focused on those variables with high correlation to quality and more 
linear behaviour:

* Alcohol
* Volatile acidity
* Sulphates (logistic)
* Chlorides

Other variables were not used to prevent redundancy, for example, citric acid, 
fixed acidity, pH and density are related to Volatile Acidity anyway.

Density is also related to alcohol. Sugar Residual and SO2 were not highly
related to quality. I included Total SO2 to improve some more the model but it's
marginal.

Even with the model its quite difficult to get a model with a strongher Rsquared
since for some quality levels there are not enough observations. That's why I
opted to collapese and merge some levels.

### Were there any interesting or surprising interactions between features?

Yes, density requires more analysis and SO2 variables as well since they show
complex relations to quality.

------

# Final Plots and Summary

### Plot One
                 3      4       5       6       7       8
Alcohol Median   9.925  10.000  9.700   10.500  11.500  12.150

```{r echo=FALSE, Plot_One}
    ggplot(aes(x = alcohol), data = pf) +
    scale_y_sqrt()+
    geom_histogram(binwidth = 0.5, color = 'darkgrey', fill = 'grey') + 
    theme(legend.position="bottom") + 
    geom_freqpoly(binwidth = 0.5, aes(color = quality_level), size = 1) +  
    scale_color_brewer(type='seq') +
    xlab('% alcohol by volume') +
    ylab('Number of observations') +
    labs(title ='Alcohol distribution and relation to quality')
```

### Description One
Starting at level 5, Alcohol follow a strong linear relation to quality. 
Levels 3 and 4 are more volatile even level 4 is bimodal.


### Plot Two
```{r echo=FALSE, Plot_Two}
  ggplot(aes(x = quality_group, y = volatile.acidity), data = pf) +
  coord_cartesian(ylim = olimits(pf$volatile.acidity)) +
  geom_boxplot(color = 'darkgrey') + 
  theme(legend.position="none") +
  labs(y='acetic acid - g/dm^3', x='quality',title='Linear relation between volatile acidity and Quality') + 
  geom_smooth(method = "lm", se=TRUE, formula = y ~x, aes(group=1))
```

### Description Two
Volatile acidity follow a linear patterns agains quality. Sadly, quality levels
3 and 8 don't have enough observations to show a better linear model. If we
collapse to 4 quality levels you can show a better behaviour.

### Plot Three
```{r echo=FALSE, Plot_Three}
  ggplot(pf, aes(x=quality_level, y=sulphates)) + 
  geom_violin(trim=FALSE)+
  scale_y_log10()+
  geom_boxplot(width=0.1, fill="white")+
  labs(x="quality", y = 'potassium sulphate - g/dm^3', title='Logistic relation between sulphates and quality')
```

### Description Three
By using conditional summaries we can see there is a logistic line to describe
quality based on sulphates with a upper limit at 0.740 g/dm3.

------

# Reflection
The model was simple in the sense that all variables were continuos. Plotting
distributions and relationships was easy with ggplot and specially ggpairs to
see multiple plots at once is just great. Using violin-plots to identify the 
impact of outlier and strange distributions is great too. But at the end one of
the main tools I used to double check my finding was the conditional summaries
table.

Once I analyzed relationship between input variables I decided to go for 4 
variables since they were related to quality and they also had strong relations
to other "secondary" variables like citric acid, pH, density, sugar residual
and fixed acidity. 

* Alcohol
* Volatile acidity
* Sulphate
* Chloride

I thouht Citric Acid would play a mayor role in my model but
at the end it didn't help that much.

Challenges:

Quality levels with just a few observations. I spent the last weeks trying to
figure it out how to deal with just a few observations at quality level 3 and 8.
Without using inference statistics I opted to grouped those categories to see
more patterns and then decide an approach.

Most variables don't have a well defined linear relationship with quality
variable so more complex analysis need to be applied to improve the model.
For example SO2 variables have concave shapes.

Removing or hidding outliers. In most cases I just used coord_cartesian to 
visually exlude them without affecting the results since I didn't had much more
information to decide it they were contamination or valid data.

Future analysis:

Quality **level 5** for chlorides, sugar residuals, sulphates
and alcohol clearly shows in violin plots the effect of outliers in that
particular group. I opted to leave the outliers but it would be interesting
to see that particular quality level in detail and fine tune some more the model.

Density is an interesting variable. It links acidity variables and alcohol/sulfur 
variables and has a great pattern to model classification since quality levels
seems to follow linear models as well. 


